The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-14 talk/vmil2012/paper.tex
rename low-level resume data into backend map

%___________________________________________________________________________
\todo{find a better name for \texttt{low-level resume data}}
\todo{mention somewhere that it is to be expected that most guards do not fail}
\section{Introduction}

@@

First a special data
structure called \emph{low-level resume data} is created. This data structure encodes the
information about where, i.e. which register or stack location, the IR-variables required to rebuild the state will be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that uses
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline the pointer to the
\emph{low-level resume data} is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \texttt{compensation code},
that is used to leave the compiled trace in case of a guard failure.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

For the different benchmarks the \texttt{low-level resume data} has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the high and low-level
resume data combined and being compressed as described before.

>>>>

%___________________________________________________________________________
\todo{mention somewhere that it is to be expected that most guards do not fail}
\section{Introduction}

@@

First a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from the IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that uses
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline the pointer to the
backend map is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \emph{compensation code},
that is used to leave the compiled trace in case of a guard failure.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{backend map} and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{backend map} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{backend map}
is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

For the different benchmarks the \texttt{backend map} has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the resume data and the backend map
combined and being compressed as described before.
